---
title: "Health and Economic Impact of Storm and Severe Weather Events"
author: ""
date: "10/19/2015"
output: html_document
---


This analysis is based on a dataset from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. The database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. The health impact is calculated using the number of fatalities and injuries. The economic impact is calculated using the data from property and crop damage.

# Data Processing

We download and unzip the storm data from the NOAA (unless those data have been downloaded and unzipped already).

```{r download}
if (! file.exists("data/StormData.csv")){
  suppressMessages(library(R.utils))
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "data/StormData.csv.bz2")
  bunzip2("data/StormData.csv.bz2", "data/StormData.csv")
}
```

Then we read the data in using read.csv and look at the number of variables and measurements (columns and rows).  All column names are then printed.

```{R read_data, cache=TRUE}
#knitr::opts_chunk$set(cache=TRUE)
stormdata <- read.csv("data/StormData.csv")
dim(stormdata)
colnames(stormdata)
```


# Results

## Health Impact

**The first question: Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?**


First let's see how many unique event types are there in the dataset
```{r event_type}
length(unique(stormdata$EVTYPE))
```

In order to evaluate the health impact we look at the numbers of fatalities and injuries. We find out which event type has the largest fatalities and injuries by averaging them over the different event types, then sorting. This method is useful for calculating the health impact per event, if we were interested in the total impact of all events of the same type, we would use sums instead of means or medians.

After sorting, we plot the first 20 event types with the hightest impact for each case.


```{r fatalities_injuries}
nr = 20 # number of events types to plot
# I did check that there are no NAs, so na.rm=true is not really necessary
meanFatalities <- tapply(stormdata$FATALITIES, stormdata$EVTYPE, mean, na.rm = TRUE, simplify=TRUE)
largestFatalities = tail(sort(meanFatalities), nr)
meanInjuries <- tapply(stormdata$INJURIES, stormdata$EVTYPE, mean, na.rm = TRUE, simplify=TRUE)
largestInjuries = tail(sort(meanInjuries), nr)
# plot using barchart from lattice, because it has the nice vertical axis with the labels
library(lattice)
par(mfrow=c(1,2))
plot1 <- barchart(largestFatalities, main="Highest health impact", xlab="fatalities")
plot2 <- barchart(largestInjuries,   main=""  , xlab="injuries")
print(plot1, position = c(0, 0, 0.5, 1), more = TRUE)
print(plot2, position = c(0.5, 0, 1, 1))
```

As we can see from the plots, Tornadoes are the ones with the highest fatalities and heat waves the ones with most injuries. We could combine fatalities and injuries with weight factors to produce a single scale for the health impact but this is left for a different study.


## Economic Impact

**The second question: Across the United States, which types of events have the greatest economic consequences?**

In order to determine the answer to this question we at property damage (PROPDMG) and crop damage (CROPDMG) but these variables must be combined with the multiplying factors using the respective exponentials in PROPDMGEXP and CROPDMGEXP, respectively. 

We print out the tables for the exponentials and then we write a function that converts the exponentials to multipliers. Then we use sapply to calculate the value of the damages for property and crops.

```{r economicImpact}
# print out the values of the exponential and their frequencies
table(stormdata$PROPDMGEXP)
table(stormdata$CROPDMGEXP)
# conversion function for EXP
multiplier_from_exponential <- function(x){
  if(x == "B" | x == "b"){ # billion
    return(1000000000)
  }else if(x == "M" | x == "m"){ # million
    return(1000000)
  }else if(x=="K" | x=="k"){ # kilo, thousand
    return(1000)
  } else if (x == "H" | x == "h") { # hecto, hundred
    return(100)
  } else if(x == ""){
    return(1)
  } else if (as.integer(x) %in% 0:8){
    return(10^(as.integer(x)))
  } else{
    # there are only a few values of the exponential which I do not understand,
    # so I am zeroing the multiplying factor for the moment.
    # This should have no considerable numerical impact
    return(0)
  }
}
propMultipliers <- sapply(stormdata$PROPDMGEXP, multiplier_from_exponential)
cropMultipliers <- sapply(stormdata$CROPDMGEXP, multiplier_from_exponential)
stormdata$propImpact <- stormdata$PROPDMG * propMultipliers
stormdata$cropImpact <- stormdata$CROPDMG * cropMultipliers
stormdata$econImpact <- stormdata$propImpact + stormdata$cropImpact
```

Now that we have the impact data in the proper units, let's find out the mean damage per event type for property and crop damage. As in the health impact case, let's plot the first 20 events with highest impact for each case.

```{r property_crop}
meanProp <- tapply(stormdata$propImpact, stormdata$EVTYPE, mean, na.rm = TRUE, simplify=TRUE)
largestProp = tail(sort(meanProp), nr)
meanCrop <- tapply(stormdata$cropImpact, stormdata$EVTYPE, mean, na.rm = TRUE, simplify=TRUE)
largestCrop = tail(sort(meanCrop), nr)
par(mfrow=c(1,3))
plot3 <- barchart(largestProp, main="Highest economic impact", xlab="property")
plot4 <- barchart(largestCrop,   main=""  , xlab="crop")
print(plot3, position = c(0, 0, 0.5, 1), more = TRUE)
print(plot4, position = c(0.5, 0, 1, 1))
```

Tornadoes are the severe weather events with the highest impact on property damage, while severe wetness has the largest impact on crop damage. Now let's also calculate the events with the largest economic impact for the total property and crops damages added together.

```{r property_total}
meanEcon <- tapply(stormdata$econImpact, stormdata$EVTYPE, mean, na.rm = TRUE, simplify=TRUE)
largestEcon = tail(sort(meanEcon), nr)
barchart(largestEcon,   main="Highest economic impact"  , xlab="total economic damage (property + crops)")
```

Since property damage is in general much higher than crop damage, the total economic damage is similar to the property damage with small changes. For example, excessive wetness, which is the highest-impact event for crops, becomes more important than hurricane opal in the total economic impact.

# Summary and Conclusions

We used data from the NOAA severe weather database in order to determine the weather events with the highest health and economic impact. From our study, tornadoes and heat waves are the events with the highest fatalities and injuries, respectively. When it comes to economic impact, tornadoes are the ones with the highest property damage, while severe wetness has the largest impact on crops. In average, the highest total economic impact comes from tornadoes.


