---
title: "Human and Economic Impact of Storm and Severe Weather Events"
author: "Dorian Kcira"
date: "10/19/2015"
output: html_document
---

# Introduction

This analysis is based on data from U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. The database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

# Data Processing


Below we download and unzip the storm data from the NOAA (unless those data have been downloaded and unzipped already).

```{r download}
if (! file.exists("data/StormData.csv")){
  suppressMessages(library(R.utils))
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "data/StormData.csv.bz2")
  bunzip2("data/StormData.csv.bz2", "data/StormData.csv")
}
```

Now let's read the data and look at the number of variables and measurements (columns and rows). Let's also print out all column names.

```{R read_data, cache=TRUE}
#knitr::opts_chunk$set(cache=TRUE)
stormdata <- read.csv("data/StormData.csv", nrows = 100)
dim(stormdata)
colnames(stormdata)
```



 str(stormdata)
'data.frame':	902297 obs. of  37 variables:
 $ STATE__   : num  1 1 1 1 1 1 1 1 1 1 ...
 $ BGN_DATE  : Factor w/ 16335 levels "10/10/1954 0:00:00",..: 6523 6523 4213 11116 1426 1426 1462 2873 3980 3980 ...
 $ BGN_TIME  : Factor w/ 3608 levels "000","0000","00:00:00 AM",..: 212 257 2645 1563 2524 3126 122 1563 3126 3126 ...
 $ TIME_ZONE : Factor w/ 22 levels "ADT","AKS","AST",..: 7 7 7 7 7 7 7 7 7 7 ...
 $ COUNTY    : num  97 3 57 89 43 77 9 123 125 57 ...
 $ COUNTYNAME: Factor w/ 29601 levels "","5NM E OF MACKINAC BRIDGE TO PRESQUE ISLE LT MI",..: 13513 1873 4598 10592 4372 10094 1973 23873 24418 4598 ...
 $ STATE     : Factor w/ 72 levels "AK","AL","AM",..: 2 2 2 2 2 2 2 2 2 2 ...
 $ EVTYPE    : Factor w/ 985 levels "?","ABNORMALLY DRY",..: 830 830 830 830 830 830 830 830 830 830 ...
 $ BGN_RANGE : num  0 0 0 0 0 0 0 0 0 0 ...
 $ BGN_AZI   : Factor w/ 35 levels "","E","Eas","EE",..: 1 1 1 1 1 1 1 1 1 1 ...
 $ BGN_LOCATI: Factor w/ 54429 levels "","?","(01R)AFB GNRY RNG AL",..: 1 1 1 1 1 1 1 1 1 1 ...
 $ END_DATE  : Factor w/ 6663 levels "","10/10/1993 0:00:00",..: 1 1 1 1 1 1 1 1 1 1 ...
 $ END_TIME  : Factor w/ 3647 levels "","?","0000",..: 1 1 1 1 1 1 1 1 1 1 ...
 $ COUNTY_END: num  0 0 0 0 0 0 0 0 0 0 ...
 $ COUNTYENDN: logi  NA NA NA NA NA NA ...
 $ END_RANGE : num  0 0 0 0 0 0 0 0 0 0 ...
 $ END_AZI   : Factor w/ 24 levels "","E","ENE","ESE",..: 1 1 1 1 1 1 1 1 1 1 ...
 $ END_LOCATI: Factor w/ 34506 levels "","(0E4)PAYSON ARPT",..: 1 1 1 1 1 1 1 1 1 1 ...
 $ LENGTH    : num  14 2 0.1 0 0 1.5 1.5 0 3.3 2.3 ...
 $ WIDTH     : num  100 150 123 100 150 177 33 33 100 100 ...
 $ F         : int  3 2 2 2 2 2 2 1 3 3 ...
 $ MAG       : num  0 0 0 0 0 0 0 0 0 0 ...
 $ FATALITIES: num  0 0 0 0 0 0 0 0 1 0 ...
 $ INJURIES  : num  15 0 2 2 2 6 1 0 14 0 ...
 $ PROPDMG   : num  25 2.5 25 2.5 2.5 2.5 2.5 2.5 25 25 ...
 $ PROPDMGEXP: Factor w/ 19 levels "","-","?","+",..: 17 17 17 17 17 17 17 17 17 17 ...
 $ CROPDMG   : num  0 0 0 0 0 0 0 0 0 0 ...
 $ CROPDMGEXP: Factor w/ 9 levels "","?","0","2",..: 1 1 1 1 1 1 1 1 1 1 ...
 $ WFO       : Factor w/ 542 levels "","2","43","9V9",..: 1 1 1 1 1 1 1 1 1 1 ...
 $ STATEOFFIC: Factor w/ 250 levels "","ALABAMA, Central",..: 1 1 1 1 1 1 1 1 1 1 ...
 $ ZONENAMES : Factor w/ 25112 levels "","                                                                                                                               "| __truncated__,..: 1 1 1 1 1 1 1 1 1 1 ...
 $ LATITUDE  : num  3040 3042 3340 3458 3412 ...
 $ LONGITUDE : num  8812 8755 8742 8626 8642 ...
 $ LATITUDE_E: num  3051 0 0 0 0 ...
 $ LONGITUDE_: num  8806 0 0 0 0 ...
 $ REMARKS   : Factor w/ 436781 levels ""," ","  ","   ",..: 1 1 1 1 1 1 1 1 1 1 ...
 $ REFNUM    : num  1 2 3 4 5 6 7 8 9 10 ...


# Results


The first question: Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

First let's see how many unique event types are there in the dataset
```{r event_type}
length(unique(stormdata$EVTYPE))
```

# Varia

R. Peng's template: http://rpubs.com/rdpeng/13396

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

   * National Weather Service Storm Data Documentation: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf 
   * National Climatic Data Center Storm Events FAQ: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf

## Assignment
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.

**Questions**

Your data analysis must address the following questions:

   1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
   2. Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

## Requirements

For this assignment you will need some specific tools

   * RStudio: You will need RStudio to publish your completed analysis document to RPubs. You can also use RStudio to edit/write your analysis.
   * knitr: You will need the knitr package in order to compile your R Markdown document and convert it to HTML

## Document Layout

   * Language: Your document should be written in English.
   * Title: Your document should have a title that briefly summarizes your data analysis
   * Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.
   * There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.
   * There should be a section titled Results in which your results are presented.
   * You may have other sections in your analysis, but Data Processing and Results are required.
   * The analysis document must have at least one figure containing a plot.
   * Your analyis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.
   * You must show all your code for the work in your analysis document. This may make the document a bit verbose, but that is okay. In general, you should ensure that echo = TRUE for every code chunk (this is the default setting in knitr).

## Publishing Your Analysis

For this assignment you will need to publish your analysis on RPubs.com. If you do not already have an account, then you will have to create a new account. After you have completed writing your analysis in RStudio, you can publish it to RPubs by doing the following:

In RStudio, make sure your R Markdown document (.Rmd) document is loaded in the editor

   1. Click the Knit HTML button in the doc toolbar to preview your document.
   1. In the preview window, click the Publish button.
   1. Once your document is published to RPubs, you should get a unique URL to that document. Make a note of this URL as you will need it to submit your assignment.

NOTE: If you are having trouble connecting with RPubs due to proxy-related or other issues, you can upload your final analysis document file as a PDF to Coursera instead.

